When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-04-03 doc/paper/vpd_et_paper.tex
finished funtional form section, and addressed reviewer and pierre's response.

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If plant response reduces ET with atmospheric
drying then soil moisture will be better conserved. This would seem a
sensible evolutionary strategy to cope with aridity. If stomata were
fully passive \citep [similar to soil pores, e.g. ][]{Or_2013},
increased atmospheric aridity would strongly reduce soil moisture
\citep{Berg_2017}. In turn, this would further increase aridity as low
soil moisture levels increase the Bowen ratio, and cause increased
temperature and atmospheric drying \citep[][]{Bouchet_1963,
  Morton_1965, Brutsaert_1999, Ozdogan_2006, Salvucci_2013,
  Gentine_2016, Berg_2016}. This however would not seem to be a
sensible strategy for plants from an evolutionary standpoint.

@@

The purpose of this study is to examine ecosystem response to
atmospheric drying, focusing on the growing season. To accomplish
this, we filter and quality control the data using a similar procedure
as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data,
  according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the
  average Gross Primary Productivity (GPP) exceeds 10\% of the
  observed 95th percentile of GPP for a given site. GPP is calculated
  using the nighttime respiration partitioning method.
\item We remove days with rain and the day following to avoid issues
  with rain interception and sensor saturation at high relative
  humidity (\cite{MEDLYN_2011}).
\end{itemize}
Additionally, as in \citet{Lin_2018}, we restrict data to the daytime,
which is identified when downwelling shortwave radiation is greater
than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W
m$^{-2}$. To reduce the chance of sensor saturation at high relative
humidity, we remove all time steps for which VPD is less than .01 kPa,
and to reduce errors at low windspeeds we remove all periods with wind
magnitudes less than 0.5 m s$^{-1}$. Timesteps with negative observed
GPP or ET are also removed, and we aggregate half hourly data to
hourly averages to reduce noise \citep{Lin_2018}. After these quality
control procedures, 434,765 upscaled hourly observations remain.

@@

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
    values are calculated, if applicable.}
  \label{definitions}
  \centering \small
  \begin{tabular}{l c c c}
    \hline
    Variable & Description & Units & Citation \\
    \hline
    $e_s$  & saturation vapor pressure & Pa  & - \\ 
    $T$  & temperature  & K & - \\
    $P$  & pressure & Pa  & - \\
    $\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
    $R_{net}$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
    $g_a$  & aerodynamic conductance & m s$^{-1}$  & \citet{Shuttleworth_2012} \\
    $\rho_a$  & air density & kg m$^{-3}$  & - \\
    $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
    $VPD$  & vapor pressure deficit & Pa  & - \\
    $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
    $g_s$  & stomatal conductance & m s$^{-1}$  & \citet{MEDLYN_2011} \\
    $g_{l-s}$  & leaf-scale stomatal conductance & mol m$^{-2}$ s$^{-1}$  & \citet{MEDLYN_2011} \\
    $R$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
    $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
    $LAI$ & leaf area index & -& - \\
    $\sigma$ & uncertainty parameter & -& - \\
    $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
    $\lambda$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
    $\Gamma$ & CO$_2$ compensation point & - & - \\
    $\Gamma^*$ & CO$_2$ compensation point without dark respiration & - & - \\
    \hline
  \end{tabular}
\end{table}

\citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$)
by combining an optimal photosynthesis theory \citep{Farquhar_1980,
  Katul_2010} with an empirical approach, which describes the
dependence of $g_s$ to VPD. This resulted in the following model for
leaf-scale stomatal conductance:

\begin{linenomath*}
  \begin{equation}
    g_{s-leaf} = g_0 + 1.6 \left(1 + \frac{g_{1-leaf}}{\sqrt{VPD}}\right) \frac{A}{c_a},
  \end{equation}
\end{linenomath*}
where $g_{1-leaf}$ is a leaf-scale ``slope'' parameter, A is the net
CO$_2$ assimilation rate, and $c_a$ is the atmospheric CO$_2$
concentration at the leaf surface. \cite{MEDLYN_2011} relate the slope
parameter ($g_{1-leaf}$) to physical parameters as:
\begin{linenomath*}
  \label{slope}
  \begin{equation}
    g_{1-leaf} = \sqrt{\frac{3 \, \Gamma^* \, \lambda}{1.6}},\footnote{Note this expression has units of of (mmol mol$^{-1}$)$^{1/2}$, but this can be converted to Pa$^{1/2}$ using the ideal gas law.}
  \end{equation}
\end{linenomath*}

@@

Thus, with this uncertainty analysis we can evaluate departure from
our theory in observations, as a departure of $\sigma$ from unity. The
variability of $\sigma$ across sites and time provides a measure of
uncertainty in our model, assumptions, as well as in the FLUXNET2015
observations themselves. The variability of $\sigma$ then propagates
through any uncertainty to our derivative of Equation \ref{et_sigma}:

@@

With Equation \ref{d_et} we have an analytical framework for ecosystem
response to atmospheric demand perturbations, with an implicit
assumption of non-extreme drought conditions associated with our
constant $uWUE$ assumption (e.g. we might expect $uWUE$ to decrease in
extreme drought). There are a few subtleties to taking the derivative
in Equation \ref{d_et}: $\Delta$ ($\frac{d e_{s}}{d T}$) and $VPD$ are
functionally related, so while taking the derivative we evaluate
$\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET}
{\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD}
\Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH}
\frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$
    fixed}}$. $RH$ and $e_s$ are assumed to be approximately
independent, which is supported by the data (not shown).

@@

Biases between the theory and observations are similiar for ENF, WSA,
CRO and SAV. At low VPD the theory and the observations match
well. However, at high VPD (upper 10-20\% of observations) the theory
is biased slightly towards postive reponse as compared to the
observations (Figure \ref{test_sign}). For ENF this slight bias could
be explained by a negative bias in $g_1$. However, for WSA, CRO, and
SAV the observations in the highest VPD bins exhibit a downturn
towards more negative response, which cannot be captured by the
functional form of the sign term (Figure
\ref{idealized_sign}). Therefore, to explain the observations one of
the varibles in the sign term must change at extreme VPD (upper
10-20\%). The most likely candidiate is $uWUE$, which we have assumed
constant to meet our goal of developing intuition for leading order
behavior, but might be expected to decrease at extremely low SWC,
which should be correlated with high VPD. The theory for EBF exhibits
similiar limiatations as for ENF, WSA, and SAV, except a larger
portion of observations are biased negative as compared to the theory
($\approx$ 35\%, Figure \ref{test_sign}). However, in general, and
specifically for non-extreme VPD, the theory matches the observations
for the tree and savannah plant types.

@@

While the above discussion shows that our theory has some limitations
when applied to some PFTs, especially for grasslands, it does well for
DBF, MF and CSH PFTs, and captures the response at non-extreme VPD for
ENF, EBF, CRO, WSA, and SAV. In general, the leading order behavior
observed in the data is captured by the theory. Departures between our
theory and observations, specifically at high VPD, could be explained
and conceptualized with shifts in $g_1$ and/or $uWUE$ due to
site-spcific plant type variability (e.g. more arid-adapted ecosystems
at more arid sites), or temporal variability for some enviornmental
conditions (e.g. decreases in $uWUE$ at extremely low soil
moisture). To focus on general behaviro and develop intuition for
PFT-scale response, we have ignored these sources of variability in
the present analysis. However, any site-specific or temporal plant
functional variability that can be conceptualized with shifts to $g_1$
and $uWUE$ can be analyzed with our framework. This opens a door to a
myriad future analysis in which plant behavior in anomalous conditions
can be explained and analyzed using Equation \ref{et}.

@@

\subsection{Limitations of theory - very dry soil moisture conditions}
In formulating our theory with Penman Monteith, we implicitly did not
account for very dry soil moisture conditions. For the majority of
environmental conditions observed at the eddy covariance sites used
here, soil conditions were not extremely dry so that we could assume a
constant uWUE and $g_1$. We posit that ecosystems will generally
optimize to host plants living in conditions which they evolved
for. However, in extreme conditions and drought scenarios soil water
content (SWC) could become the limiting factor for ET response to VPD,
which our theory does not account for. In addition, low soil moisture
conditions themselves increase VPD through land-atmosphere feedback
\citep[][]{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006,
  Salvucci_2013, Gentine_2016, Berg_2016}.

@@

To summarize, our theory is limited by its inability to account for
soil water impacts on land surface response, and feedbacks between SWC
and VPD. Fortunately, for most PFTs SWC's effect on ET is relatively
rare (<30\% of observations) and does not manifest itself in the
majority of observations and bulk statistics. However, for GRA SWC has
a strong effect decreasing water use efficiency for the majority of
the observations. Soil moisture effects explain the deficiencies of
our theory in Section \ref{testing}, particularly for GRA. By
conceptualizing SWC effects as a change in $uWUE$ (and/or $g_1$), it
will be possible for future analysis to explore the importance of soil
moisture on plant response to VPD.

\subsection{Functional form of ET dependence on VPD and its relation
  to VPD dependence}

Th theory described in Section \ref{sign_term} indicates that for a
given $uWUE$ and $g_1$, the ET dependence on VPD should be concave
upward. In other words, there should be some local minima in ET at a
critical VPD$_{crit}$, assuming the scaling term (e.g. aerodynamic
conductance and $\Delta$) is held fixed. This result warrants futher
investigation, because to our knowledge no one has derived the
theoretical ecosystem-scale relationship between ET and VPD while
controlling for enivironmental conditiions. In particular, from
personal communication, there is an apparent lack of consensus over
whether the shape of the ET-VPD curve should be concave upward (our
result) or concave downward. Given that understanding the ET-VPD
relationship of the one-way plant response is fundamental to
hypothesizing about any feedbacks between the land surface and the
atmopshere, we analyze why our derived ET-VPD relationship is concave
upward, particularly with respect to the exponent of VPD depedence in
$uWUE$ and the Medlyn unified stomatal conductace model.

There is a theoretical basis for the square root VPD dependence in
both the stomatal conducatance model and $uWUE$ based on the
assumption that stomata behave to maximize carbon gain while
minimizing water loss, which observations also generally support
\citep{Lloyd_1991, MEDLYN_2011, Lin_2015, Zhou_2014, Zhou_2015,
  Medlyn_2017}. However, purely empiriacle results which fit the
exponent of the VPD dependence have shown that it may vary from 1/2,
suggesting that stomata, as well as ecosystem-scale quantities based
on stomata theroy, may not always function optimally \citep{Zhou_2015,
  Lin_2018}. Specifically with regards to $uWUE$, one would not expect
that this ecosystem scale WUE quantity will responde to VPD exactly
analgalously to stomata. Direct soil evaporation's contributions to ET
should shift the exponential dependence of VPD to be greater than
1/2. In other words, direct soil evaporation causes WUE to decrease
more strongly with increasing VPD, especially in well-watered
conditions. \citet{Zhou_2015}'s results corroborate this: they found a
mean imprically fit exponential VPD dependence of 0.55 istead of 0.5
for Ameriflux sites. Given that these recent results on the
relatioship between VPD, GPP, and ET form the backbone of our analysis
and are what allowed us to derive an explicit expression for ET for
the first time (Equation \ref{et}), we will analyze if and how
assumptions about the exonent of the VPD dependence impacts the shape
of the ET-VPD dependence. By introducing $n$ and $m$ we can free our
stomatal conductance model from assumptions about VPD dependence:
\begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_1}{VPD^m}\right) \frac{*WUE \; ET}{c_a \; VPD^n},
    \label{m_n}
  \end{equation}
\end{linenomath*}
where:
\[*WUE = \frac{GPP}{ET}VPD^n.\] To determine how the exponent $n$ and
$m$ alter the shape of the ET-VPD dependence we find the roots of the
second deriviative of ET, using Equation \ref{m_n} for stomatal
conductance, with respect to VPD: \input{d2_solutions.tex}

>>>>

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If ecosystem plant response reduces ET with atmospheric
drying then soil moisture will be better conserved. This would seem a
sensible evolutionary strategy to cope with aridity. If stomata were
fully passive \citep [similar to soil pores, e.g. ][]{Or_2013},
increased atmospheric aridity would strongly reduce soil moisture
\citep{Berg_2017}. In turn, this would further increase aridity as low
soil moisture levels increase the Bowen ratio, and cause increased
temperature and atmospheric drying \citep[][]{Bouchet_1963,
  Morton_1965, Brutsaert_1999, Ozdogan_2006, Salvucci_2013,
  Gentine_2016, Berg_2016}. This however would not seem to be a
sensible strategy for plants from an evolutionary standpoint. One may
argue that increases in ET with increasing VPD could increase the
liklihood of precipitation \citep[e.g.,][]{Findell_2011}. However,
increases in ET are not a gaurantee
of increased likelhood of precipitation, which also dependes on the
environemtnal conditions \citep[][]{Gentine_2013}. Futhermore,  any
increases in precipitation are likely to be non-local, such that plants giving up
water to the atmosphere are not gauranteed to reap the benefits of
water returned to the atmosphere. Using this logic, from an ecoysystem
evolutionary perspective, water stored in soil is worth much more than
the \textit{chance} of water returned as precipitation.

@@

The purpose of this study is to examine ecosystem response to
atmospheric drying, focusing on the growing season. To accomplish
this, we filter and quality control the data using a similar procedure
as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data,
  according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the
  average Gross Primary Productivity (GPP) exceeds 10\% of the
  observed 95th percentile of GPP for a given site. GPP is calculated
  using the nighttime respiration partitioning method.
\item We remove days with rain and the day following to avoid issues
  with rain interception and sensor saturation at high relative
  humidity (\cite{Medlyn_2017}).
\end{itemize}
Additionally, as in \citet{Lin_2018}, we restrict data to the daytime,
which is identified when downwelling shortwave radiation is greater
than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W
m$^{-2}$. To reduce the chance of sensor saturation at high relative
humidity, we remove all time steps for which VPD is less than .01 kPa,
and to reduce errors at low windspeeds we remove all periods with wind
magnitudes less than 0.5 m s$^{-1}$. Timesteps with negative observed
GPP or ET are also removed, and we aggregate half hourly data to
hourly averages to reduce noise \citep{Lin_2018}. After these quality
control procedures, 434,765 upscaled hourly observations remain.

@@

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
    values are calculated, if applicable.}
  \label{definitions}
  \centering \small
  \begin{tabular}{l c c c}
    \hline
    Variable & Description & Units & Citation \\
    \hline
    $e_s$  & saturation vapor pressure & Pa  & - \\ 
    $T$  & temperature  & K & - \\
    $P$  & pressure & Pa  & - \\
    $\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
    $R_{net}$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
    $g_a$  & aerodynamic conductance & m s$^{-1}$  & \citet{Shuttleworth_2012} \\
    $\rho_a$  & air density & kg m$^{-3}$  & - \\
    $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
    $VPD$  & vapor pressure deficit & Pa  & - \\
    $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
    $g_{s-leaf}$  & leaf-scale stomatal conductance & m s$^{-1}$  & \citet{MEDLYN_2011} \\
    $g_{s}$  &  stomatal conductance & m s$^{-1}$
                                   & \citet{Medlyn_2017} \\
    $g_{1-leaf}$  & leaf-scale slope parameter & Pa$^{0.5}$
                                   & \citet{MEDLYN_2011} \\
    $g_{1}$  & ecoysystem-scale slope parameter & Pa$^{0.5}$ & \citet{MEDLYN_2017} \\
    $R$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
    $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
    $\sigma$ & uncertainty parameter & -& - \\
    $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
    $\lambda$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
    $\Gamma$ & CO$_2$ compensation point & - & - \\
    $\Gamma^*$ & CO$_2$ compensation point without dark respiration & - & - \\
    \hline
  \end{tabular}
\end{table}

\citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$)
by combining an optimal photosynthesis theory \citep{Cowan_1977} with an empirical approach, which describes the
dependence of $g_s$ to VPD. This resulted in the following model for
leaf-scale stomatal conductance:

\begin{linenomath*}
  \begin{equation}
    g_{s-leaf} = g_0 + 1.6 \left(1 +
      \frac{g_{1-leaf}}{\sqrt{VPD}}\right) \frac{A}{c_a},
    \label{leaf_medlyn}
  \end{equation}
\end{linenomath*}
where $g_{1-leaf}$ is a leaf-scale ``slope'' parameter, A is the net
CO$_2$ assimilation rate, and $c_a$ is the atmospheric CO$_2$
concentration at the leaf surface. \cite{MEDLYN_2011} relate the slope
parameter ($g_{1-leaf}$) to physical parameters as:
\begin{linenomath*}
  \label{slope}
  \begin{equation}
    g_{1-leaf} = \sqrt{\frac{3 \, \Gamma^* \, \lambda}{1.6}},\footnote{Note this expression has units of of (mmol mol$^{-1}$)$^{1/2}$, but this can be converted to Pa$^{1/2}$ using the ideal gas law.}
  \end{equation}
\end{linenomath*}

@@

Thus, with this uncertainty analysis we can evaluate departure from
our theory in observations, as a departure of $\sigma$ from unity. The
variability of $\sigma$ across sites and time provides a measure of
uncertainty in our model, assumptions, as well as in the FLUXNET2015
observations themselves. The variability of $\sigma$ then propagates
through any uncertainty to our partial derivative of Equation
\ref{et_sigma} with respect to VPD:

@@

With Equation \ref{d_et} we have an analytical framework for ecosystem
response to atmospheric demand perturbations with environmental
conditions held fixed. There are a few subtleties to taking the
derivative in Equation \ref{d_et}: $\Delta$ ($\frac{d e_{s}}{d T}$)
and $VPD$ are functionally related, so while taking the derivative we
evaluate
$\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET}
{\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD}
\Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH}
\frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$
    fixed}}$. $RH$ and $e_s$ are assumed to be approximately
independent, which is supported by the data (not shown).

This derivation relied either implicitly or explicitly on several
assumptions. First, the underlying theory used to derive Equations
\ref{medlyn}, \ref{leaf_medlyn} rely on a implicit assumptions that
$ET=g_s \cdot VPD$; physically this implies that leaves are perfectly
coupled to the atmosphere. In reality, for some conditions and plant
types the leaves can become decoupled from the boundary layer
\citep{De_2017, Medlyn_2017}. Given our focus on the growing season
which is usually characterized by relatively high insolation, we would
expect the surface to be well coupled. Additionally, by following
\citet{Medlyn_2017}'s method for calcualting eco-system scale $g_1$,
we have stronger confidence that surface coupling is represented
within our framework, even if it is not represented in the underlying
theory used to develop $g_s$. Additionally, the formulation of $uWUE$
in \citet{Zhou_2014} and \citet{Zhou_2015} as well as our use of
\citet{Medlyn_2017}'s stomatal conductance model assumes that that
direct soil evaporation contributions to ET remian relatively
small. This should be more true druing the growing season than other
times; however some plant types allow for systematically larger
contributions of ``E'' in ET, particularly those with less dense
canopies and smaller aboslute amounts of transpiration. We therefore
might expect that the theory will be most applicable to forest PFTs,
which will be most strongly coupled to the boundary layer due to
larger surface roughness, and will also generally have the highest
ratios of transpiration to evaporation. Finally, for the goal of
developing PFT-wide intuition, we assume that $g_1$ and $uWUE$ are
constant in space and time. Both quantities have been shown to be well
conserved over space and time within a PFT relative to inter-PFT
variability. However, we would expect them to vary somehwat with
environmental conditions including soil water content temperature, and
with intra-PFT plant-specific characteristics like wood density for
tree PFTs \citep{Lin_2015}. Given these strong assumptions made with
the goal of understanding broad, leading order plant behaviors, we
take the extra care of including and analyzing uncertainty (Section
\ref{testing}) and examing sensitivity to the exponent of the VPD
response derived using assumptions of $ET=g_s \cdot VPD$ (Section
\ref{functional_form}).

@@

Biases between the theory and observations are similiar for ENF, WSA,
CRO and SAV. At low VPD the theory and the observations match
well. However, at high VPD (upper 10-20\% of observations) the theory
is biased slightly towards postive reponse as compared to the
observations (Figure \ref{test_sign}). For ENF this slight bias could
be explained by a negative bias in $g_1$. However, for WSA, CRO, and
SAV the observations in the highest VPD bins exhibit a downturn
towards more negative response, which cannot be captured by the
functional form of the sign term (Figure
\ref{idealized_sign}). Therefore, to explain the observations one of
the varibles in the sign term must change at extreme VPD (upper
10-20\%). The most likely candidiate is $uWUE$, which we have assumed
constant to meet our goal of developing intuition for leading order
behavior, but might be expected to decrease at extremely low SWC,
which should be correlated with high VPD. The theory for EBF exhibits
similiar limiatations as for ENF, WSA, and SAV, except a larger
portion of observations are biased negative as compared to the theory
($\approx$ 35\%, Figure \ref{test_sign}). However, in general, and
specifically for non-extreme VPD (VPD < $\approx$70-90th percentiles), the theory matches the observations
for the tree and savannah plant types.

@@

While the above discussion shows that our theory has some limitations
when applied to some PFTs, especially for grasslands, it does well for
DBF, MF and CSH PFTs, and captures the response at non-extreme VPD
(VPD < $\approx$ 70th-90th percentiles) for
ENF, EBF, CRO, WSA, and SAV. In general, the leading order behavior
observed in the data is captured by the theory. Departures between our
theory and observations, specifically at high VPD, could be explained
and conceptualized with shifts in $g_1$ and/or $uWUE$ due to
site-spcific plant type variability (e.g. more arid-adapted ecosystems
at more arid sites), or temporal variability for some enviornmental
conditions (e.g. decreases in $uWUE$ at extremely low soil
moisture). To focus on general behaviro and develop intuition for
PFT-scale response, we have ignored these sources of variability in
the present analysis. However, any site-specific or temporal plant
functional variability that can be conceptualized with shifts to $g_1$
and $uWUE$ can be analyzed with our framework. This opens a door to a
myriad future analysis in which plant behavior in anomalous conditions
can be explained and analyzed using Equation \ref{et}.

@@

\subsection{Limitations of theory - very dry soil moisture conditions}
\label{swc_section}
In formulating our theory with Penman Monteith, we implicitly did not
account for very dry soil moisture conditions. For the majority of
environmental conditions observed at the eddy covariance sites used
here, soil conditions were not extremely dry so that we could assume a
constant uWUE and $g_1$. We posit that ecosystems will generally
optimize to host plants living in conditions which they evolved
for. However, in extreme conditions and drought scenarios soil water
content (SWC) could become the limiting factor for ET response to VPD,
which our theory does not account for. In addition, low soil moisture
conditions themselves increase VPD through land-atmosphere feedback
\citep[][]{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006,
  Salvucci_2013, Gentine_2016, Berg_2016}.

@@

To summarize, our theory is limited by its inability to account for
soil water impacts on land surface response, and feedbacks between SWC
and VPD. Fortunately, for most PFTs SWC's effect on ET is relatively
rare (<30\% of observations) and does not manifest itself in the
majority of observations and bulk statistics. This result is
consistent with However, for GRA SWC has
a strong effect decreasing water use efficiency for the majority of
the observations. Soil moisture effects explain the deficiencies of
our theory in Section \ref{testing}, particularly for GRA. By
conceptualizing SWC effects as a change in $uWUE$ (and/or $g_1$), it
will be possible for future analysis to explore the importance of soil
moisture on plant response to VPD.

\subsection{Functional form of ET dependence on VPD and its relation
  to VPD dependence}
\label{functional_form}

The theory described in Section \ref{sign_term} indicates that for a
given $uWUE$ and $g_1$, the ET dependence on VPD should be concave
upward. In other words, there should be some local minima in ET at a
critical VPD$_{crit}$, assuming the scaling and plant terms  (e.g. aerodynamic
conductance, $\Delta$, $g_1$ and $uWUE$) are held fixed. This result warrants futher
investigation, because to our knowledge no one has derived the
theoretical ecosystem-scale relationship between ET and VPD while
controlling for enivironmental conditiions. In particular, from
personal communication, there is an apparent lack of consensus over
whether the shape of the ET-VPD curve should be concave upward (our
result) or concave downward. Given that understanding the ET-VPD
relationship of the one-way plant response is fundamental to
hypothesizing about any feedbacks between the land surface and the
atmopshere, we analyze why our derived ET-VPD relationship is concave
upward, particularly with respect to the exponent of VPD depedence in
$uWUE$ and the Medlyn unified stomatal conductace model.

There is a theoretical basis for the square root VPD dependence in
both the stomatal conducatance model and $uWUE$ based on the
assumption that stomata behave to maximize carbon gain while
minimizing water loss, which observations also generally support
\citep{Lloyd_1991, MEDLYN_2011, Lin_2015, Zhou_2014, Zhou_2015,
  Medlyn_2017}. However, some purely empiriacle results which fit the
exponent of the VPD dependence to data have shown that it may vary slightly
from 1/2, suggesting that stomata, as well as ecosystem-scale
quantities based on stomata theroy, may not always function optimally
\citep{Zhou_2015, Lin_2018}. Specifically with regards to $uWUE$, one
would not expect that this ecosystem scale WUE quantity will responde
to VPD exactly analgalously to stomata. Direct soil evaporation's
contributions to ET should shift the exponent of the VPD dependence,
especially for conditions of low GPP when we would expect a
systematically larger portion of direct soil evaporation contributions
to ET, because we would also expect lower amounts of
transpiration. \citet{Zhou_2015}'s results corroborate this: they
found a mean imprically fit exponential VPD dependence of 0.55,
varying slightly from the theoretically optimal value of of 0.5 for
Ameriflux sites. Interpreting \citet{Lin_2018}'s results, which also
show variance in the empiracle exponent of the VPD dependence of the
stomatal conductance model, is more difficult as \citet{Lin_2018} does
not handle GPP/A dependence of stomatal conductance in a directly
analgoalous manner to the optimal theory in \citet{MEDLYN_2011} and
\citet{Medlyn_2017}. Regardless, given that these recent results
\citep{MEDLYN_2011, Zhou_2014, Zhou_2015, Medlyn_2017} on the
relatioship between VPD, GPP, and ET \citep{MEDLYN_2011, Zhou_2014,
  Zhou_2015, Medlyn_2017} form the backbone of our analysis and are
what allowed us to derive an explicit ET expression for the first
time (Equation \ref{et}), we will analyze if and how assumptions about
the exponent of the VPD dependence impacts the shape of the ET-VPD
dependence. By introducing $n$ and $m$ we can free our stomatal
conductance model from assumptions about VPD dependence:
\begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_*}{VPD^m}\right) \frac{*WUE \; ET}{c_a \; VPD^n},
    \label{m_n}
  \end{equation}
\end{linenomath*}
where:
\[*WUE = \frac{GPP}{ET}VPD^n,\] and $g_*$ is a generic slope parameter
of units $VPD^m$. To determine how the exponent $n$ and $m$ alter the
shape of the ET-VPD dependence we find the roots of the second
deriviative of ET, using Equation \ref{m_n} for stomatal conductance
($g_s$, with respect to VPD: \input{d2_solutions.tex}. With this
result we've detemined a family of curves seperating concave up from
concave down ET solutions (Figure \ref{concave}). These curves are
only functions of the exponent of the VPD depedence and a quantity we
call non-dimensional VPD ($VPD^m/g*$). Several important
relations reveal themselves from Equation \ref{curves}:

\begin{figure}
  \centering
  \centerline{\includegraphics[width=1\textwidth]{./concave.pdf}}
  \caption{ Solutions corresponding to inflection points between
    concave up and concave down ET-VPD curves (Equation \ref{curves})
    for 3 specific scenarios. Solutions are defined in terms of a
    non-dimensional VPD ($VPD^m/g_*$), but to aide physical
    interpretation the horizontal axis is additionally provided interms of dimensionalized VPD assuming $m=1/2$ and
    $g_*=110\; Pa^{1/2}$ (average of all PFT $g_1$). The vertical axis
    has a different interpretation depending on the solution
    curve. For the blue line ($m$ varying), it correspondes to $m$,
    for the orange line ($n$ varying) it correspondes to $n$, and for
    the green line it correspondes to the value of both $n$ and $m$
    ($n=m$). Regions of the parameter space that corresponde to
    concave up and concave down results are labeled.}
  \label{concave}
\end{figure}

\begin{enumerate}
  \item For optimal behavior (n, m = 1/2) the ET-VPD curve will be
    concave up regardless of the magnitude of the plant constants
    $g_1$ and $uWUE$. Therefore, the general concave up nature of our
    results, given an assumption of optimal behavior, is insensitive
    to plant type.
  \item For all physically possible exponents of VPD dependence ($n,
    m$), whether the solution is concave up or concave down does not
    depend on $uWUE$.
  \item In general, increasing the exponent of VPD dependence
    increases the liklihood of a concave up result. Additionally, as
    the exponent of VPD dependence increases from the optimum value of
    1/2, whether the curve is concave upward or concave downward
    becomes a function of the plant specific slope parameter $g_*$,
    through non-dimensional VPD ($VPD^m/g_*$). Because the exponent of
    the VPD dependencies is capable of altering the fundamental shape
    of ET-VPD dependence, future research investment in understanding
    the exact VPD depedence of stomatal conductance, and further
    reconciliation of empiracle and theoretical stomatal and
    ecosostyem behavior should be prioritized.
\end{enumerate}
However, while it is possible that in the future some other form of
VPD dependence is derived, at present \cite{MEDLYN_2011} and
\cite{Zhou_2014} firmly established n=m=1/2 as the most likely
candidiate given current theory and empiracle data. Additionally, we
argue that a concave up result matches physical intuition more than a
concave down result. Plants must maintain nutrient and sugar transport
through the phloem and xylem. To accomplish this, stomata must remain
slightly open \citep{De_2013, Nikinmaa_2013, Ryan_2014}. Futhermore,
even if complete stomatal closure were possible, cuticular water loss
and [at the ecosystem-scale] direct soil evaporation are still sources
of ET which increase with VPD, independent of stomatal
closure. Therefore, in the limit as VPD becomes large and we assume
plants are exercising all strategies to reduce ET, any futher increase
in VPD should result in an increase in ET through cuticular water loss
and/or direct soil evaporation. This inevitable transition from
conditions where stomata responde strongly to VPD to conditions where
stomata response is asymptoting towards full closure would cause a
concave up ET-VPD curve, which is matched by the theory. In short,
plant response becomes more limited as VPD increases, while
atmospheric demand monotonically increases with VPD, leading to the
intuitive result that atmospheric demand dominates plant response when
atmospheric deman is high.

This analysis allow us to understand the theoretical shape of the ET
response to VPD with environmental conditions held fixed which would
be difficult to disentagle purely from flux observations, given the
relatively fast time scale of plant response and the non-stationarity
of [solar forced] environmental conditions  over the relatively
coarse (half hourly) flux estimates (which is required to obtain
robust eddy-covariance statistics). This fundamental understanding of
ET response to VPD with environmental conditoins held fixed can be
built upon with future work examing how changes in VPD and enviornment
conditions (e.g. soil water storage) feedback upon one another. In the
soil water storage example, over very long time scales extremely high
VPD perturbations coupled with no precipitation could result in
decreases in soil water storage such that water becomes limiting. This
could be represented by an extension of our framework in which uWUE is
allowed to decrease with decreasing SWC, as observed in Secion
\ref{swc_section}. Here, we focus our results by assuming constant
PFT-wide conditions to build baseline intuition for ET-VPD
depedence. For most PFTs, the theory with plant function held fixed
matches the leading order behaviro of the observations where plant
function varies (Section \ref{testing}).
